Disease Knowledge Transfer for Alzheimer's Variants

### 2018-03-02 paper/paper.tex
m

@@

% hypothetical model has specific ordering of modality biomarkers -> ordering or spatial biomarkers (different dimension) -> spatial differences in different dementias -> network vulnerability hypothesis.
A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline. While this hypothetical model proposes a modality-specific ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities. However, the spatial ordering of abnormality differs in other Alzheimer's variants. For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected. The network vulnerability theory by Seeley et al. \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration. There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}. 
% * <alexandra.young@ucl.ac.uk> 2018-03-02T19:01:16.243Z:
% 
% > The network vulnerability theory by Seeley et al. \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration. There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}. 
% The relationship between these two statements is unclear from the way its written
% 
% ^.

@@

\subsection{Methods}

For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where correlations within each unit are disease-agnostic (e.g. biomarkers corresponding to the same ROI).  Each biomarker is assigned a-priori to a specific functional unit, which corresponds to a brain region, say temporal lobe (Fig 1., bottom).  For example, biomarkers assigned to the temporal unit could be: amyloid in temporal lobe, tau in temporal lobe, MRI atrophy in temporal lobe and certain memory tests. Using a disease progression model of our choice, for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects. Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.

@@

Variable names:
\begin{itemize}
 \item $\Lambda$ = set of \emph{functional units}, i.e. \{temporal, parietal, occipital, ..\}. Biomarkers belonging to the same functional unit are asusmed to have disease-agnostic correlations.
 \item $y_{ijk}$ = measurement in subject $i$, visit $j$, biomarker $k$
%  \item $d_{ijl}$ = dysfunctionality score in subject $i$, visit $j$ in \emph{functional unit} $l$, where $l \in \Lambda$
%  \item $s_{ij}$ = disease progression score in subject $i$, visit $j$
 \item $\theta_k$ = parameters of trajectory for biomarker $k$
 \item $\lambda^l$ = parameters of trajectory for functional unit $l$, where $l \in \Lambda$
 \item $\psi$ : \{1, ..., K\} $ \rightarrow Lambda$ is a function that maps beach biomarker to its corresponding functional unit
 \item $\beta_i$: time shift parameter for subject $i$ (disease specific)
 \item $\beta_i^{l}$: time shift parameter for subject $i$ used in functional unit $l$, where $l \in \Lambda$
 \item $\Omega$: set ${(i,j,k)}$ of measurements available from every subject $i$, visit $j$ in biomarker $k$
\end{itemize}

@@

Fig. \ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit}, along with samples from the model posterior and aligned subject data. The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. Fig. \ref{fig:occipUnit} shows the progression of dysfunctionality scores for (a) typical AD and (b) PCA. In typical AD, we see that temporal dysfunction becomes abnormal

@@

% data fitting within one unit
\begin{figure}[H]
\includegraphics[width=\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
\caption{Estimated biomarker trajectories in the Occipital Unit. Subject data from ADNI and our local cohort are also shown. The X-axis represents, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject. Red lines represent samples from the trajectory posterior. The Y-axis represents biomarker values which have been normalised to the [0,1] range.}
\label{fig:occipUnit}
\end{figure}

% PCA vs tAD disease space
\begin{figure}[H]
\begin{subfigure}{0.47\textwidth}
\centering
typical AD\\
\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
\caption{}
\end{subfigure}
\begin{subfigure}{0.47\textwidth}
\centering
PCA\\
\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
\caption{}
\end{subfigure}
\caption{Progression of dysfunctionality scores for (a) typical AD and (b) PCA. In typical AD, we notice that the temporal scores are most affected after the disease }
\label{fig:pcaTadDisSpace}
\end{figure}

>>>>

% write background in abstract. motivation. take out

@@

% hypothetical model has specific ordering of modality biomarkers -> ordering or spatial biomarkers (different dimension) -> spatial differences in different dementias -> network vulnerability hypothesis.
A hypothetical model of disease progression has been previously published by \cite{jack2010hypothetical}, which proposes that the first biomarkers to become abnormal are measures of amyloid beta aggregation, followed by tau abnormalities, hypometabolism, structural MRI-based measures and finally cognitive decline. While this hypothetical model proposes a inter-modality ordering of biomarkers in typical AD, it has also been observed that within the same modality, biomarker measurements have a spatial sequence of abnormality that correlates with Braak stages: hippocampal volumes and entorhinal measures become abnormal first, followed by other structures within the temporal lobe, followed by parietal and frontal abnormalities. However, the spatial ordering of abnormality differs in other Alzheimer's variants. For example, in Posterior Cortical Atrophy, posterior regions such as the occipital lobe and superior parietal regions have been shown to become affected. More generally, the network vulnerability theory by Seeley et al. \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration. Overall, there is a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}. 
% * <alexandra.young@ucl.ac.uk> 2018-03-02T19:01:16.243Z:
% 
% > The network vulnerability theory by Seeley et al. \cite{seeley2009neurodegenerative} suggests that damage to specific underlying connectivity networks leads to spatially distinct patterns of neurodegeneration. There is nevertheless a great degree of heterogeneity within these diseases, which led some to suggest that subjects with different Alzheimer's variants lie on a continuum of phenotypical variation \cite{crutch2012posterior}. 
% The relationship between these two statements is unclear from the way its written
% 
% ^.

@@

\subsection{Methods}
% TODO: this was copy-pasted from a draft 2 weeks ago. needs re-writing!
For modelling disease-agnostic correlations, we assume biomarkers are grouped into \emph{functional units}, where correlations within each unit are disease-agnostic (e.g. biomarkers corresponding to the same ROI).  Each biomarker is assigned a-priori to a specific functional unit, which corresponds to a brain region, say temporal lobe (Fig 1., bottom).  For example, biomarkers assigned to the temporal unit could be: amyloid in temporal lobe, tau in temporal lobe, MRI atrophy in temporal lobe and certain memory tests. Using an arbitrary disease progression model (e.g. CITE), for each region-specific functional unit we can model the correlation of biomarkers within that functional unit (Figure 1, bottom), which allows us to reconstruct a region- or unit-specific \emph{dysfunction} progression axis which can be used for staging subjects. Finally, we express the disease specific correlations (Figure 1, top) not directly with each biomarkers, but via the dysfunction scores estimated within each functional unit.

@@

Variable names:
\begin{itemize}
 \item $\Lambda$ = set of \emph{functional units}, i.e. \{temporal, parietal, occipital, ..\}. Biomarkers belonging to the same functional unit are asusmed to have disease-agnostic correlations.
 \item $y_{ijk}$ = measurement in subject $i$, visit $j$, biomarker $k$
%  \item $d_{ijl}$ = dysfunctionality score in subject $i$, visit $j$ in \emph{functional unit} $l$, where $l \in \Lambda$
%  \item $s_{ij}$ = disease progression score in subject $i$, visit $j$
 \item $\theta_k$ = parameters of trajectory for biomarker $k$
 \item $\lambda^l$ = parameters of trajectory for functional unit $l$, where $l \in \Lambda$
 \item $\psi$ : \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps beach biomarker to its corresponding functional unit
 \item $\beta_i$: time shift parameter for subject $i$ (disease specific)
 \item $\beta_i^{l}$: time shift parameter for subject $i$ used in functional unit $l$, where $l \in \Lambda$
 \item $\Omega$: set ${(i,j,k)}$ of measurements available from every subject $i$, visit $j$ in biomarker $k$
\end{itemize}

@@

% Fig1: biomarker traj. over dysfunction scores in one functional unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Fig. \ref{fig:occipUnit} shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data. The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range. The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. Fig. \ref{fig:occipUnit} shows the progression of dysfunctionality scores over the disease stage for (a) typical AD and (b) PCA (disease ). In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C. Baron, Neuroimage, 2001). In Fig. \ref{fig:PCAtrajByModality} we plot the inferred biomarker trajectories for PCA across the five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451. The trajectories

@@

% data fitting within one unit
\begin{figure}
\includegraphics[width=\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
\caption{Estimated biomarker trajectories in the Occipital Unit. Subject data from ADNI and our local cohort are also shown. The X-axis represents, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject. Red lines represent samples from the trajectory posterior. The Y-axis measures biomarker values (normalised).}
\label{fig:occipUnit}
\end{figure}

% PCA vs tAD disease space
\begin{figure}
\begin{subfigure}{0.47\textwidth}
\centering
typical AD\\
\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
\caption{}
\end{subfigure}
\begin{subfigure}{0.47\textwidth}
\centering
PCA\\
\includegraphics[width=\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
\caption{}
\end{subfigure}
\caption{Progression of dysfunctionality scores for (a) typical AD and (b) PCA. In typical AD, we notice that the temporal scores are most affected after the disease }
\label{fig:pcaTadDisSpace}
\end{figure}
